Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-21 ms.tex
Suggestions from Dan Noble
Thanks @daniel1noble!

\title{Versioned data: why it’s needed and how it can be achieved (easily and cheaply)}
\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
\affiliation{
$^1$ Imperial College, London, United Kingdom\\
$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
University of New South Wales, Sydney, NSW 2052, Australia\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
}
\date{}

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.  There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.  This future system would be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently.  That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

@@

One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. This is enabled via the distributed and decentralised nature of \texttt{git} repositories. Mltiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
% We have constant drama at work because the different spreadsheet
% writers we use across the platforms quote files inconsistently so
% every save is a complete rewrite of the file.  It's totally
% impossible to do any collaborative editing....
 or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy. For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project. This process is called submitting a ``Pull request''. One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
 % This section conflates git and github more than is useful.  It
 % might be worth setting up github as the explicit platform here or
 % dialing back the language a bit.  GitLab and bitbucket have their
 % own "merge request" feature (I think).  It's also github that makes
 % repositories _discoverable_ which is a requirement for getting
 % outside collaborators in.  Oddly the decentralised nature of git is
 % not actually that useful in the github model because github is so
 % strongly centralised.  Will and I did some true decentralised git
 % collaboration on a plane to nescent once with a bare git repo on a
 % thumb drive.

@@

A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}. Ideally, users should be able to also able access the all versions of the database programmatically. Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.  To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system. We have written a \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to her data with (very) minimal computational skills. We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive. When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.

@@

We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}). However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}. Indeed, we would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.

@@

The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.  The current structure of scientific research is very distributed, with research groups across the world working on the same question.  Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{figures/Figure1.pdf}
\caption{Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
\textbf{a)} 3 numbers indicate type of change.
\textbf{b)} A typical release stream.}
\label{fig:semantic}
\end{figure}

>>>>

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
\affiliation{
$^1$ Imperial College, London, United Kingdom\\
$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
University of New South Wales, Sydney, NSW 2052, Australia\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
}
\date{}

@@

% So, git doesn't actually work this way, though everyone seems to
% think that it does.  Git stores *content* and then later tries to
% de-duplicate it.  It then displays changes as line-by-line diffs
% (but you can change this to display the changes other ways; see
% --color-words --ignore-all-space for example.  The problem is that
% the algorithms that git uses are optimised for typical sizes of code
% and not for data.  Small data is actually totally fine to store with
% git.  The other issue that turns up is if the files are _binary_
% then an arbitrarily small change in the underlying data can result
% in an arbitrarily large change in the binary representation, which
% prevents deduplication.
%
% Git-LFS tries to step around this by storing large files separately
% from git's general machinery.  Git-annex did the same.  But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with.  I _always_ have to look up how to compare two versions across
% branches.  And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.
Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”. Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach. From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate. They just don’t feel like its worth doing.
There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.  This future system would be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently.  That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.

@@

One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. This is enabled via the distributed and decentralised nature of \texttt{git} repositories. Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
% We have constant drama at work because the different spreadsheet
% writers we use across the platforms quote files inconsistently so
% every save is a complete rewrite of the file.  It's totally
% impossible to do any collaborative editing....
 or, when needed, flag where there are conflicts that need to be resolved. This makes collaboration easy. For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project. This process is called submitting a ``Pull request''. One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
 % This section conflates git and github more than is useful.  It
 % might be worth setting up github as the explicit platform here or
 % dialing back the language a bit.  GitLab and bitbucket have their
 % own "merge request" feature (I think).  It's also github that makes
 % repositories _discoverable_ which is a requirement for getting
 % outside collaborators in.  Oddly the decentralised nature of git is
 % not actually that useful in the github model because github is so
 % strongly centralised.  Will and I did some true decentralised git
 % collaboration on a plane to nescent once with a bare git repo on a
 % thumb drive.

@@

A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}. Ideally, users should be able to access all versions of the database programmatically. Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.  To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system. We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset. Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills. We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}. \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”. You may want to discuss this in greater depth as I don’t see how this is possible.
When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline). If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g. \texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded. This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.

@@

We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}. We would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.

@@

The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.
% Reword
The current structure of scientific research is very distributed, with research groups across the world working on the same question.  Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.  As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=\linewidth]{figures/Figure1.pdf}
\caption{
% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
\textbf{a)} 3 numbers indicate type of change.
\textbf{b)} A typical release stream.}
\label{fig:semantic}
\end{figure}
